Engineering spin squeezing in a 3D optical lattice with interacting spin-orbit-coupled fermions

### 2019-05-31 soc_squeezing/squeezing_paper/main.tex
fixed minor error in squeezing paper added two references on characterizing inter-atomic interactions added bibliography for squeezing paper

% The understanding and application of few-body quantum mechanics and equilibrium statistical mechanics were integral to the rapid growth in technology throughout the 20th century.
A major frontier of contemporary physics is the understanding of non-equilibrium behaviors of many-body quantum systems, and the application of these behaviors toward the development of novel quantum technologies with untapped capabilities\cite{eisert2015quantum}.
To this end, ultracold atomic, molecular, and optical systems are ideal platforms for studying unexplored regimes of many-body physics due to their clean preparation and readout, high controllability, and long coherence times\cite{bloch2008manybody, gross2017quantum}.
The exquisite capabilities of these systems have pushed the frontiers of metrology, quantum simulation, and quantum information science.

Optical lattice clocks in particular have seen some of the most impressive developments in recent years, reaching record levels of precision ($\sim 3\times 10^{-19}$)\cite{campbell2017fermidegenerate, marti2018imaging} and accuracy ($\sim 1\times 10^{-18}$)\cite{bloom2014optical, mcgrew2018atomic}.
These advancements required important breakthroughs, including the capability to cool and trap fermionic alkaline-earth atoms in spin-insensitive potentials\cite{takamoto2003spectroscopy, barber2006direct, ye2008quantum}; the development of ultra-coherent lasers\cite{kessler2012sub40mhzlinewidth, cole2013tenfold, matei2017mu} to fully exploit an ultra-narrow clock transition\cite{ludlow2015optical}; and, more recently, the preparation of a quantum degenerate gas in a three-dimensional (3D) optical lattice\cite{campbell2017fermidegenerate, marti2018imaging, goban2018emergence}.
Nonetheless, all improvements in sensing capabilities to date have been based on single-particle control of internal atomic degrees of freedom.
Such strategies will eventually have diminishing returns due to practical difficulties in (i) suppressing decoherence from external (motional) degrees of freedom, and (ii) interrogating more particles without additional systematic errors from interactions\cite{martin2013quantum, ludlow2015optical, marti2018imaging}.

@@

An external laser with Rabi frequency $\Omega$ and wavenumber $k_{\t{L}}$ along the tunneling axis resonantly couples atoms' internal states through the Hamiltonian $\hat H_{\t{laser}}/\hbar = \sum_j \Omega e^{ik_{\t{L}}x_j} \hat c_{j,\up}^\dag \hat c_{j,\dn} + \t{h.c.}$, where $\hat c_{j\alpha}$ is a fermionic annihilation operator for an atom on site $j$ with internal state $\alpha\in\set{\up,\dn}$ and $x_j$ is the position of site $j$.
This laser imprints a position-dependent phase that equates to a momentum kick $k_{\t{L}}$ when an atom changes internal states by absorbing or emitting a photon, thereby generating spin-orbit coupling\cite{wall2016synthetic, livi2016synthetic}.
After absorbing the position dependence of the laser Hamiltonian into fermionic operators through the gauge transformation $\hat c_{j,\up}\to e^{ik_{\t{L}} x_j}\hat c_{j,\up}$, which makes $\hat H_{\t{laser}}$ spatially homogeneous, the atoms are well-described in the tight-binding limit by the Fermi-Hubbard Hamiltonian\cite{esslinger2010fermihubbard}
\begin{align}
  \hat H_{\text{FH}}^{(\phi)}/\hbar
  &= -J \sum_j \p{e^{i\phi} \hat c_{j,\up}^\dag \hat c_{j+1,\up}
  + \hat c_{j,\dn}^\dag \hat c_{j+1,\dn} + \text{h.c.}} \notag \\
  &\qquad + U \sum_j \hat n_{j,\up} \hat n_{j,\dn},
\label{eq:FermiHubbard}
\end{align}
where $J$ is the nearest-neighbor tunneling rate; the SOC angle $\phi\equiv k_{\t{L}}a$ determines the phase gained by spin-up atoms upon tunneling from site $j$ to site $j+1$ (in the gauge-transformed frame) with lattice spacing $a=x_{j+1}-x_j$; $U$ is the on-site interaction energy of two atoms; and $\hat n_{j\alpha}\equiv\hat c_{j\alpha}^\dag \hat c_{j\alpha}$ is a number operator.

>>>>

A major frontier of contemporary physics is the understanding of non-equilibrium behaviors of many-body quantum systems, and the application of these behaviors toward the development of novel quantum technologies with untapped capabilities\cite{eisert2015quantum}.
To this end, ultracold atomic, molecular, and optical systems are ideal platforms for studying unexplored regimes of many-body physics due to their clean preparation and readout, high controllability, and long coherence times\cite{bloch2008manybody, gross2017quantum}.
The exquisite capabilities of these systems have pushed the frontiers of metrology, quantum simulation, and quantum information science.

Optical lattice clocks in particular have seen some of the most impressive developments in recent years, reaching record levels of precision ($\sim 3\times 10^{-19}$)\cite{campbell2017fermidegenerate, marti2018imaging} and accuracy ($\sim 1\times 10^{-18}$)\cite{bloom2014optical, mcgrew2018atomic}.
These advancements required important breakthroughs, including the capability to cool and trap fermionic alkaline-earth atoms in spin-insensitive potentials\cite{takamoto2003spectroscopy, barber2006direct, ye2008quantum}; the development of ultra-coherent lasers\cite{kessler2012sub40mhzlinewidth, cole2013tenfold, matei2017mu} to fully exploit an ultra-narrow clock transition\cite{ludlow2015optical}; the detailed characterization of inter-atomic interactions\cite{scazza2014observation, cappellini2014direct}; and, more recently, the preparation of a quantum degenerate gas in a three-dimensional (3D) optical lattice\cite{campbell2017fermidegenerate, marti2018imaging, goban2018emergence}.
Nonetheless, all improvements in sensing capabilities to date have been based on single-particle control of internal atomic degrees of freedom.
Such strategies will eventually have diminishing returns due to practical difficulties in (i) suppressing decoherence from external (motional) degrees of freedom, and (ii) interrogating more particles without additional systematic errors from interactions\cite{martin2013quantum, ludlow2015optical, marti2018imaging}.

@@

An external laser with Rabi frequency $\Omega$ and wavenumber $k_{\t{L}}$ along the tunneling axis resonantly couples atoms' internal states through the Hamiltonian $\hat H_{\t{laser}}/\hbar = \sum_j \Omega e^{ik_{\t{L}}x_j} \hat c_{j,\up}^\dag \hat c_{j,\dn} + \t{h.c.}$, where $\hat c_{j\alpha}$ is a fermionic annihilation operator for an atom on site $j$ with internal state $\alpha\in\set{\up,\dn}$ and $x_j$ is the position of site $j$.
This laser imprints a position-dependent phase that equates to a momentum kick $k_{\t{L}}$ when an atom changes internal states by absorbing or emitting a photon, thereby generating spin-orbit coupling\cite{wall2016synthetic, livi2016synthetic}.
After absorbing the position dependence of the laser Hamiltonian into fermionic operators through the gauge transformation $\hat c_{j,\up}\to e^{ik_{\t{L}} x_j}\hat c_{j,\up}$, which makes $\hat H_{\t{laser}}$ spatially homogeneous, the atoms are well-described in the tight-binding limit by the Fermi-Hubbard Hamiltonian\cite{esslinger2010fermihubbard}
\begin{align}
  \hat H_{\t{FH}}^{(\phi)}/\hbar
  &= -J \sum_j \p{e^{i\phi} \hat c_{j,\up}^\dag \hat c_{j+1,\up}
  + \hat c_{j,\dn}^\dag \hat c_{j+1,\dn} + \text{h.c.}} \notag \\
  &\qquad + U \sum_j \hat n_{j,\up} \hat n_{j,\dn},
\label{eq:FermiHubbard}
\end{align}
where $J$ is the nearest-neighbor tunneling rate; the SOC angle $\phi\equiv k_{\t{L}}a$ determines the phase gained by spin-up atoms upon tunneling from site $j$ to site $j+1$ (in the gauge-transformed frame) with lattice spacing $a=x_{j+1}-x_j$; $U$ is the on-site interaction energy of two atoms; and $\hat n_{j\alpha}\equiv\hat c_{j\alpha}^\dag \hat c_{j\alpha}$ is a number operator.
